A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-05-07 paper/paper.tex
paper: change to MSCS class; rewrite introduction

%% -*- latex-command: pdflatex -shell-escape -*-
\documentclass[10pt]{art.cls/art}

\usepackage{multirow, float, tabularx, environ, booktabs, setspace, makecell, caption, art.cls/joinargs, colortbl, multicol, rotating, flushend}

\makeatletter
\@namedef{ver@lineno.sty}{9999/12/31}
\@namedef{opt@lineno.sty}{}
\makeatother
\usepackage{minted}

@@

% Simplicial and Cubical sets
\newcommand{\DeltaHat}{\ensuremath{\hat{\boldsymbol{\Delta}}}}
\newcommand{\DeltaPlus}{\ensuremath{\boldsymbol{\Delta}_+}}
\newcommand{\Cube}{\ensuremath{\boldsymbol{\square}}}
\newcommand{\CSet}{\ensuremath{\mathsf{Set}_{\boldsymbol{\square}}}}

@@

@@

% The eqntable environment, displaying the various
\newcolumntype{Y}{>{\centering\arraybackslash}X}
\NewEnviron{onecoltbl}[1]{
  \begin{table}[H]
    \small
    \begin{tabularx}{\linewidth}{@{}>{$}l<{$} >{$}c<{$} >{$}c<{$} >{$}Y<{$} @{}}
      \toprule
      \BODY
      \bottomrule
    \end{tabularx}
    \caption{#1}
  \end{table}
}

\NewEnviron{eqntable}[1]{
  \begin{table*}[!t]
    \small
    \begin{tabularx}{\linewidth}{@{}>{$}l<{$} >{$}c<{$} >{$}c<{$} >{$}Y<{$} @{}}
      \toprule
      \BODY
      \bottomrule
    \end{tabularx}
    \caption{#1}
  \end{table*}
}

@@

\title{A parametricity-based formalization of semi-simplicial and semi-cubical sets}
\author{\IEEEauthorblockN{Hugo Herbelin \\ and Ramkumar Ramachandra}\\\IEEEauthorblockA{Université Paris Cité, Inria, CNRS, IRIF, Paris\\\texttt{Hugo.Herbelin@inria.fr} and \texttt{r@artagnon.com}}}

\begin{document}
\maketitle
\begin{abstract}
  Constructions such as semi-simplicial and semi-cubical sets can be defined in the ``usual way'' as presheaves over respectively, the semi-simplex or semi-cube category, which we call \emph{fibered} definitions, but also defined like in e.g. Voevodsky~\cite{voevodsky12} or in previous work~\cite{herbelin15}, as a dependently-typed construction, which we call \emph{indexed}.

  This paper describes a uniform indexed characterization of both augmented semi-simplicial and semi-cubical sets arising respectively as unary and binary iterated parametricity-based constructions.

  Additionally, our construction is fully formalized in Coq's dependent type theory.
\end{abstract}

@@

@@

@@

\begin{definition}[$\DeltaPlus$]
  \begin{figure*}[!t]
    \begin{align*}
      \Obj(\DeltaPlus) & := \Nat                                                                     \\
      \Hom(p, n)       & := \{l \in [0, \kstar]^n \mid \text{number of $\kstar$ in $l = p$}\}        \\
      g \circ f        & :=
      \begin{cases}
        f                  & \text{if $g = []$}                                                        \\
        0 :: (g' \circ f)  & \text{if $g = (0 :: g')$}                                                 \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = 0$ or $\kstar$} \\
      \end{cases} \\
      \id              & := [\kstar, \ldots, \kstar] \text{ $n$ times for $\id \in \Hom(n, n)$}
    \end{align*}
    \caption{Definition of \DeltaPlus}\label{fig:deltaplus}
  \end{figure*}

@@

\begin{definition}[\Cube]
  The definition of $\Cube$ is shown in Fig.~\ref{fig:cube}. The symbols $L$ and $R$ indicate opposite faces of a cube.
  \begin{figure*}[!t]
    \begin{align*}
      \Obj(\Cube) & := \Nat                                                                                \\
      \Hom(p, n)  & := \{l \in [L, R, \kstar]^n \mid \text{number of $\kstar$ in $l = p$}\}                \\
      g \circ f   & :=
      \begin{cases}
        f                  & \text{if $g = []$}                                                              \\
        a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $a = L$ or $R$}                          \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = L$, $R$, or $\kstar$} \\
      \end{cases} \\
      \id         & := [\kstar, \ldots, \kstar] \text{ $n$ times}
    \end{align*}
    \caption{Definition of \Cube}\label{fig:cube}
  \end{figure*}

@@

\begin{figure*}[!t]
  \begin{equation*}
    \begin{array}{ll}
      X_0                                : \underbrace{\unittype}_{\framep[0][0]}                                                    \rightarrow \U \\
      X_1                                : \underbrace{\Sigma \unitpoint: \unittype. \underbrace{\left(
      \begin{array}{c}
          \underbrace{X_0(\unitpoint)}_{\painting[0][0]}
          \\ \times \\
          \underbrace{X_0(\unitpoint)}_{\painting[0][0]}
        \end{array}\right)}_{\layer[1][0]}}_{\framep[1][1]} \rightarrow \U                                                                            \\
      X_2                                : \underbrace{\Sigma a: \underbrace{\left(\Sigma \unitpoint: \unittype. \underbrace{\left(
          \begin{array}{c}
            \underbrace{\Sigma b: \left(
            \hspace{-0.4em}\begin{array}{c}
                             X_0(\unitpoint)
                             \\ \times \\
                             X_0(\unitpoint)
                           \end{array}\hspace{-0.4em}
            \right).\, \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][L]}}_{\painting[1][1]}}_{\painting[1][0]}
            \\ \times \\
            \underbrace{\Sigma b: \left(
            \hspace{-0.4em}\begin{array}{c}
                             X_0(\unitpoint)
                             \\ \times \\
                             X_0(\unitpoint)
                           \end{array}\hspace{-0.4em}
            \right).\, \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][R]}}_{\painting[1][1]}}_{\painting[1][0]}
          \end{array}
          \right)}_{\layer[2][0]}\right)}_{\framep[2][1]}.\underbrace{\left(
        \begin{array}{c}
          \underbrace{X_1 \underbrace{\left(a.\hd, \left(
              \hspace{-0.4em}\begin{array}{c}
                               a.\tl.L.\hd.L, \\
                               a.\tl.R.\hd.L
                             \end{array}\hspace{-0.4em}
              \right)\right)}_{\restrf[2][1][L]}}_{\painting[1][1]}
          \\ \times \\
          \underbrace{X_1 \underbrace{\left(a.\hd, \left(
              \hspace{-0.4em}\begin{array}{c}
                               a.\tl.L.\hd.R, \\
                               a.\tl.R.\hd.R
                             \end{array}\hspace{-0.4em}
              \right)\right)}_{\restrf[2][1][R]}}_{\painting[1][1]}
        \end{array}
        \right)}_{\layer[2][1]}}_{\framep[2][2]}
      \rightarrow \U                                                                                                                                \\
      \ldots
    \end{array}
  \end{equation*}
  \caption{Intuition for formal construction ($\nu = 2$)}\label{fig:intuition}
\end{figure*}

@@

\begin{figure*}[!t]
  \begin{align*}
    \framep[n][p](D)      & \defeq & \Sigma d: (\ldots (\Sigma \unitpoint: \unittype.\, \layer[n][0](D)(\unitpoint)) \ldots).\, \layer[n][p](D)(d)                                \\
    \painting[n][p](D)(d) & \defeq & \Sigma l_p: \layer[n][p](D)(d).\, (\ldots (\Sigma l_n: \layer[n][n - 1](D)(d, l_p, \ldots, l_{n - 1}).\, D.\tl(d, l_p, \ldots, l_n)) \ldots) \\
                          &        & \text{where } (d, l_p, \ldots, l_q) \text{ abbreviates } ((\ldots(d, l_p), \ldots), l_q)
  \end{align*}
  \caption{Refinement of definition of $\framep$ and $\painting$\label{fig:refinement}}
\end{figure*}

@@

\renewcommand*{\lab}{fulltab:coind}
\input{tab-coind.tex}
\renewcommand*{\lab}{fulltab:core}
\input{tab-core.tex}
\renewcommand*{\lab}{fulltab:frames}
\input{tab-frames.tex}

\begin{sidewaystable*}
  \renewcommand*{\lab}{fulltab:faces}
  \input{tab-faces.tex}
  \vspace{5em}
  \renewcommand*{\lab}{fulltab:coh}
  \input{tab-coh.tex}
\end{sidewaystable*}

@@

\bibliographystyle{IEEEtran}
\bibliography{paper}

>>>>

\documentclass{msc}

\usepackage{amsmath, amssymb, mathrsfs, url, bookmark, tikz, tikz-cd, mathpazo, enumitem, xargs, multirow, float, tabularx, environ, booktabs, setspace, makecell, caption, colortbl, pdflscape, flushend, minted, art.cls/colorpal, art.cls/ct, art.cls/sset, art.cls/lim, art.cls/joinargs}

@@

% Simplicial and Cubical sets
\newcommand{\DeltaHat}{\ensuremath{\hat{\boldsymbol{\Delta}}}}
\newcommand{\DeltaPlus}{\ensuremath{\boldsymbol{\Delta}_+}}
\newcommand{\Cube}{\ensuremath{\boldsymbol{{\Box}}}}
\newcommand{\CSet}{\ensuremath{\mathsf{Set}_{\boldsymbol{\Box}}}}

@@

% Examples
\newtheorem{example}[therm]{Example}

@@

% The eqntable environment
\newcolumntype{Y}{>{\centering\arraybackslash}X}
\NewEnviron{eqntable}[1]{
  \begin{table}[H]
    \small
    \begin{tabularx}{\linewidth}{
      @{}
      >{$}l<{$}
      >{$}c<{$}
      >{$}c<{$}
      >{$}Y<{$}
      @{}}
      \toprule
      \BODY
      \bottomrule
    \end{tabularx}
    \caption{#1}
  \end{table}
}

@@

% Use lettered footnotes
\renewcommand{\thefootnote}{\alph{footnote}}

\begin{document}
\title{A parametricity-based formalization of semi-simplicial and semi-cubical sets}
\begin{authgrp}
  \author{Hugo Herbelin}
  \affiliation{Université Paris Cité, Inria, CNRS, IRIF, Paris
    \email{Hugo.Herbelin@inria.fr}}
\end{authgrp}
\begin{abstract}
  Constructions such as semi-simplicial and semi-cubical sets can be defined in the ``usual way'' as presheaves over respectively, the semi-simplex or semi-cube category, which we call \emph{fibered} definitions, but also defined like in e.g. Voevodsky~\cite{voevodsky12} or in previous work~\cite{herbelin15}, as a dependently-typed construction, which we call \emph{indexed}.

  On another side, it is known that semi-simplicial and semi-cubical sets are related to iterated Reynolds's parametricity, respectively in its unary and binary variants\dots

  We exploit this correspondence to develop a uniform indexed definition of both augmented semi-simplicial and semi-cubical sets, which is additionally fully formalized in Coq's dependent type theory. Beside the interest in the construction itself, we expect it to eventually serve as models of type theory preserving more definitional equalities than presheaf models do.
\end{abstract}

@@

\subsection{Fibered vs indexed presentation of semi-simplicial and semi-cubical sets}

A dependent set can be represented in two ways: as a family indexed over a set or as a set together with a map telling for each element on which index it depends on. In the latter case, the set associated to a given index is the fiber (i.e. the inverse image) of this index under the map. The two presentations are equivalent, what can be formally stated as what we call the fibered/indexed correspondence\footnote{At a discussion on the Zulip discussion channel for the HoTT community, alternative names fibration/family and unbundled/bundled were proposed, among others. In our experience, the opposition fibered/indexed is intuitively understood by many speakers. In particular, it echoes to the Grothendieck construction of fibered categories from indexed categories, so that some people also calls this correspondence the Grothendieck construction for dummies}:
$$(\Sigma S: \U. (S \rightarrow T)) \simeq (T \rightarrow \U)$$

While a presheaf, i.e. a family of sets with maps between them according to an index category, can be seen as living on the indexed side of the correspondence, with discrete fibration on the other side of the correspondence, there are situations where the presheaf can be seen as living on the fibered side of the correspondence.
Indeed, when the index category of the presheaf is direct, i.e. with a well-founded collection of morphisms, as is the case of semi-simplicial and semi-cubical sets, each set serving as domain of a map can be turned into a family dependent, i.e. indexed, on the codomain of the map, and, since the collection of maps is well-founded, this can be iterated, eventually producing a collection of families which we call the indexed presentation of the presheaf over a direct category\dots

For instance, in the case of a semi-cubical set, presented with $2n$ face maps from the set of $n$-cubes to the set of $(n-1)$-cubes, the presheaf approach corresponds to a family of sets fibered the ones over the others as follows (up to some coherence conditions):

\begin{equation*}
  \begin{tikzcd}
    X_0: \U & X_1: \U \arrow[l, "\delta^L" description, shift left=2] \arrow[l, "\delta^R" description, shift right=2] & X_2: \U \arrow[l, "\delta^{L\kstar}" description, shift left=6] \arrow[l, "\delta^{R\kstar}" description, shift left=2] \arrow[l, "\delta^{\kstar L}" description, shift right=2] \arrow[l, "\delta^{\kstar R}" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}

On the other side, the indexed presentation of the same presheaf is a stratified sequence of families recursively indexed on elements of families of lower rank, taking into account the coherence conditions to prevent duplications:
\begin{align*}
  X_0                                                                                      & \mbox{ a set of points}                                       \\
  (X_1)_{a,b\in X_0}                                                                       & \mbox{ a set of lines between any two points}                 \\
  (X_2)_{a,b,c,d\in X_0,p\in (X_1)_{ab},q\in (X_1)_{bc},r \in (X_1)_{ac}, s\in (X_1)_{cd}} & \mbox{ a set of squares over any 4 lines lying over 4 points} \\
  \ldots
\end{align*}

The idea for such indexed presentation of presheaves over a direct category was introduced at the Univalent Foundations year in the context of defining semi-simplicial types~\cite{...}. A couple of constructions were proposed since then. One, by Voevodsky, relies on the presentation of semi-simplicial sets as a presheaf over injective monotone finite maps (see \cite[Section]{Herbelin14} for a presentation). Another, by the first author, and fully formalized in Coq, relies on the presentation of semi-simplicial sets as a presheaf over face maps~\cite{Herbelin14}. Another one by Park and Luo, and formalized in Plastic, works generically for any Reedy category with only upwards maps~\cite{ParkLuo}.

The indexed definition of a presheaf over a direct category is technically more involvedthanthepresheafdefinition as it requires hard-wiringinthestructurethe dependencies betweenelements of the sets of the presheaf, including the coherence condition between these dependencies, such as taking the $i$-th face of the $j$-th face of a $n$-simplex is the same as taking the $(j-1)$-th face of the $i$-th face (when $j>i$). On the other side, defining a concrete presheaf in indexedform only requires to provide the family since the burden of defining maps and showing the coherence conditions over them is has already been integrate what is not the case in the presheaf presentation.

\subsection{Reynolds' parametricity and its unary and binary variants}

@@

\subsection{Iterated-parametricity-based constructio}

@@

\begin{definition}[$\DeltaPlus$]
  \begin{figure}[H]
    \begin{align*}
      \Obj(\DeltaPlus) & := \Nat                                                                     \\
      \Hom(p, n)       & := \{l \in [0, \kstar]^n \mid \text{number of $\kstar$ in $l = p$}\}        \\
      g \circ f        & :=
      \begin{cases}
        f                  & \text{if $g = []$}                                                        \\
        0 :: (g' \circ f)  & \text{if $g = (0 :: g')$}                                                 \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = 0$ or $\kstar$} \\
      \end{cases} \\
      \id              & := [\kstar, \ldots, \kstar] \text{ $n$ times for $\id \in \Hom(n, n)$}
    \end{align*}
    \caption{Definition of \DeltaPlus}\label{fig:deltaplus}
  \end{figure}

@@

\begin{definition}[\Cube]
  The definition of $\Cube$ is shown in Fig.~\ref{fig:cube}. The symbols $L$ and $R$ indicate opposite faces of a cube.
  \begin{figure}[H]
    \begin{align*}
      \Obj(\Cube) & := \Nat                                                                                \\
      \Hom(p, n)  & := \{l \in [L, R, \kstar]^n \mid \text{number of $\kstar$ in $l = p$}\}                \\
      g \circ f   & :=
      \begin{cases}
        f                  & \text{if $g = []$}                                                              \\
        a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $a = L$ or $R$}                          \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = L$, $R$, or $\kstar$} \\
      \end{cases} \\
      \id         & := [\kstar, \ldots, \kstar] \text{ $n$ times}
    \end{align*}
    \caption{Definition of \Cube}\label{fig:cube}
  \end{figure}

@@

\begin{figure}[H]
  \begin{equation*}
    \begin{array}{ll}
      X_0                                : \underbrace{\unittype}_{\framep[0][0]}                                                    \rightarrow \U \\
      X_1                                : \underbrace{\Sigma \unitpoint: \unittype. \underbrace{\left(
      \begin{array}{c}
          \underbrace{X_0(\unitpoint)}_{\painting[0][0]}
          \\ \times \\
          \underbrace{X_0(\unitpoint)}_{\painting[0][0]}
        \end{array}\right)}_{\layer[1][0]}}_{\framep[1][1]} \rightarrow \U                                                                            \\
      X_2                                : \underbrace{\Sigma a: \underbrace{\left(\Sigma \unitpoint: \unittype. \underbrace{\left(
          \begin{array}{c}
            \underbrace{\Sigma b: \left(
            \hspace{-0.4em}\begin{array}{c}
                             X_0(\unitpoint)
                             \\ \times \\
                             X_0(\unitpoint)
                           \end{array}\hspace{-0.4em}
            \right).\, \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][L]}}_{\painting[1][1]}}_{\painting[1][0]}
            \\ \times \\
            \underbrace{\Sigma b: \left(
            \hspace{-0.4em}\begin{array}{c}
                             X_0(\unitpoint)
                             \\ \times \\
                             X_0(\unitpoint)
                           \end{array}\hspace{-0.4em}
            \right).\, \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][R]}}_{\painting[1][1]}}_{\painting[1][0]}
          \end{array}
          \right)}_{\layer[2][0]}\right)}_{\framep[2][1]}.\underbrace{\left(
        \begin{array}{c}
          \underbrace{X_1 \underbrace{\left(a.\hd, \left(
              \hspace{-0.4em}\begin{array}{c}
                               a.\tl.L.\hd.L, \\
                               a.\tl.R.\hd.L
                             \end{array}\hspace{-0.4em}
              \right)\right)}_{\restrf[2][1][L]}}_{\painting[1][1]}
          \\ \times \\
          \underbrace{X_1 \underbrace{\left(a.\hd, \left(
              \hspace{-0.4em}\begin{array}{c}
                               a.\tl.L.\hd.R, \\
                               a.\tl.R.\hd.R
                             \end{array}\hspace{-0.4em}
              \right)\right)}_{\restrf[2][1][R]}}_{\painting[1][1]}
        \end{array}
        \right)}_{\layer[2][1]}}_{\framep[2][2]}
      \rightarrow \U                                                                                                                                \\
      \ldots
    \end{array}
  \end{equation*}
  \caption{Intuition for formal construction ($\nu = 2$)}\label{fig:intuition}
\end{figure}

@@

\begin{figure}[H]
  \begin{align*}
    \framep[n][p](D)      & \defeq & \Sigma d: (\ldots (\Sigma \unitpoint: \unittype.\, \layer[n][0](D)(\unitpoint)) \ldots).\, \layer[n][p](D)(d)                                \\
    \painting[n][p](D)(d) & \defeq & \Sigma l_p: \layer[n][p](D)(d).\, (\ldots (\Sigma l_n: \layer[n][n - 1](D)(d, l_p, \ldots, l_{n - 1}).\, D.\tl(d, l_p, \ldots, l_n)) \ldots) \\
                          &        & \text{where } (d, l_p, \ldots, l_q) \text{ abbreviates } ((\ldots(d, l_p), \ldots), l_q)
  \end{align*}
  \caption{Refinement of definition of $\framep$ and $\painting$\label{fig:refinement}}
\end{figure}

@@

\renewcommand*{\lab}{fulltab:coind}
\input{tab-coind.tex}
\renewcommand*{\lab}{fulltab:core}
\input{tab-core.tex}
\renewcommand*{\lab}{fulltab:frames}
\input{tab-frames.tex}
\renewcommand*{\lab}{fulltab:faces}
\input{tab-faces.tex}

\begin{landscape}
  \renewcommand*{\lab}{fulltab:coh}
  \input{tab-coh.tex}
\end{landscape}

@@

Another direction for future work would also be to generically construct the indexed form of any presentation of presheaves over a direct category and to show the equivalence between the two presentations.
\bibliographystyle{msclike}
\bibliography{paper}
